---
title: "Stage SCUC Classified Copies"
subtitle: "Copy reviewed source files into 06, 07, and 08 without altering originals"
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
execute:
  warning: false
  message: false
  error: false
editor: source
---

## Safety model

This workbook consumes the manifest created by `00_scuc_dataset_match_audit.qmd`. It only copies files; it never moves, renames, or deletes originals. Relative source paths are preserved beneath each classification folder, avoiding collisions and retaining provenance.

The default is a dry run. Before enabling copies, review `04 project inventory outputs/scuc_dataset_match_manifest.csv`, especially every row assigned to `08 datasets prior to pre-canonical set matches`.

## 1. Settings and manifest validation

```{r}
#| label: settings

recovery_root <- fs::path_abs(getwd())

search_root <- fs::path_abs(fs::path(recovery_root, ".."))

manifest_path <- fs::path(recovery_root, "04 project inventory outputs", "scuc_dataset_match_manifest.csv")

copy_log_path <- fs::path(recovery_root, "04 project inventory outputs", "scuc_dataset_copy_log.csv")

dry_run <- FALSE

overwrite_existing <- FALSE

if (!fs::file_exists(manifest_path)) {
  stop("Run 00_scuc_dataset_match_audit.qmd first. Missing manifest: ", manifest_path)
}

manifest <- readr::read_csv(manifest_path, show_col_types = FALSE, progress = FALSE)

required <- c("classification", "source_path", "byte_sha256", "copy_eligible", "proposed_destination")

missing_columns <- setdiff(required, names(manifest))
if (length(missing_columns)) stop("Manifest is missing: ", paste(missing_columns, collapse = ", "))

allowed_classes <- c(
  "06 canonical data set matches", "07 pre-canonical set matches",
  "08 datasets prior to pre-canonical set matches"
)
if (any(!manifest$classification %in% allowed_classes)) stop("Manifest contains an unknown classification")

copy_plan <- manifest |>
  dplyr::filter(copy_eligible) |>
  dplyr::mutate(
    source_now_exists = fs::file_exists(source_path),
    source_hash_now = vapply(source_path, function(p) {
      if (fs::file_exists(p)) digest::digest(file = p, algo = "sha256") else NA_character_
    }, character(1)),
    source_unchanged_since_audit = source_now_exists & source_hash_now == byte_sha256,
    destination_is_inside_class_folder = vapply(seq_len(dplyr::n()), function(i) {
      startsWith(fs::path_abs(proposed_destination[i]),
                 paste0(fs::path(recovery_root, classification[i]), "/"))
    }, logical(1)),
    destination_exists_before = fs::file_exists(proposed_destination),
    action = dplyr::case_when(
      !source_now_exists ~ "SKIP: source missing",
      !source_unchanged_since_audit ~ "SKIP: source changed since audit",
      !destination_is_inside_class_folder ~ "SKIP: unsafe destination",
      destination_exists_before & !overwrite_existing ~ "SKIP: destination exists",
      dry_run ~ "DRY RUN: would copy",
      TRUE ~ "COPY"
    )
  )

copy_plan |>
  dplyr::count(classification, action, sort = TRUE)

```




## 2. Copy eligible files

Set `dry_run <- FALSE` only after reviewing the plan above. Re-running is safe by default because existing destinations are skipped.

```{r}
#| label: copy

if (!dry_run) {
  for (i in which(copy_plan$action == "COPY")) {
    fs::dir_create(fs::path_dir(copy_plan$proposed_destination[i]), recurse = TRUE)
    copy_error <- tryCatch({
      fs::file_copy(copy_plan$source_path[i], copy_plan$proposed_destination[i],
                    overwrite = overwrite_existing)
      NA_character_
    }, error = function(e) conditionMessage(e))
    copy_plan$action[i] <- if (is.na(copy_error)) {
      "COPIED; pending hash verification"
    } else {
      paste0("COPY FAILED: ", copy_error)
    }
  }
}

copy_plan <- copy_plan |>
  dplyr::mutate(
    destination_exists_after = fs::file_exists(proposed_destination),
    destination_hash_after = vapply(proposed_destination, function(p) {
      if (fs::file_exists(p)) digest::digest(file = p, algo = "sha256") else NA_character_
    }, character(1)),
    verified_copy = destination_exists_after & destination_hash_after == byte_sha256,
    action = dplyr::case_when(
      verified_copy & grepl("^COPIED", action) ~ "COPIED AND VERIFIED",
      !verified_copy & grepl("^COPIED", action) ~ "COPY COMPLETED BUT HASH VERIFICATION FAILED",
      TRUE ~ action
    ),
    run_time = format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"),
    was_dry_run = dry_run
  )

readr::write_csv(copy_plan, copy_log_path, na = "")
copy_plan |>
  dplyr::count(classification, action, verified_copy, sort = TRUE)

```




## After staging

Use the staged folders as a comparison/archive workspace:

- In **06**, replace use of each duplicate with the manifest’s named canonical reference.
- In **07**, preserve one documented pre-canonical predecessor where historical provenance matters.
- Review **08** manually. This workflow intentionally does not delete it or assert that unmatched means disposable.
- Delete or archive original files only as a separate, deliberate operation after the copy log shows `verified_copy = TRUE` and you have a backup.

