---
title: "SCUC Dependency Scan and Cleanup Plan"
subtitle: "Find active references before archiving any original dataset"
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
execute:
  warning: false
  message: false
  error: false
editor: source
---

## Purpose and safety boundary

This workbook scans source-code and project-configuration files under `/Users/D/R Working Directory` for references to the 39 original datasets in the approved match manifest. It produces a dependency report and a conservative cleanup plan.

It does **not** edit code, move files, archive files, or delete anything.

Dependency confidence is deliberately conservative:

- **certain**: the code contains the original absolute or relative path;
- **likely**: the original filename appears in code inside the same top-level project;
- **ambiguous**: only the filename appears elsewhere and may refer to another copy.

## 1. Settings

```{r}
#| label: settings

recovery_root <- fs::path_abs(getwd())
search_root <- fs::path_abs(fs::path(recovery_root, ".."))
manifest_path <- fs::path(recovery_root, "04 project inventory outputs", "scuc_dataset_match_manifest.csv")
copy_log_path <- fs::path(recovery_root, "04 project inventory outputs", "scuc_dataset_copy_log.csv")
report_dir <- fs::path(recovery_root, "05 dependency and cleanup reports")

source_extensions <- c(
  "r", "rmd", "qmd", "md", "yml", "yaml", "json", "toml", "txt", "sh"
)

# Owner-confirmed datasets that are intentionally separate from the canonical
# SCUC Snapshot wide/long series.
separate_dataset_filenames <- c(
  "TEST SCUC 2004 TEA TAKS Data Grades 3-11.csv",
  "scuc-tea-5yr-snapshot-df.xlsx",
  "scuc-tea-5yr-2017-2021 snapshots.xlsx"
)

excluded_directory_names <- c(
  ".git", ".Rproj.user", ".quarto", "_freeze", "_site", "renv", "packrat",
  "node_modules", "library", "history_archive", "history_review",
  "SCUC_discovery_review", "04 project inventory outputs",
  "05 dependency and cleanup reports", "06 canonical data set matches",
  "07 pre-canonical set matches", "08 datasets prior to pre-canonical set matches"
)

if (!fs::file_exists(manifest_path)) stop("Missing approved manifest: ", manifest_path)
if (!fs::file_exists(copy_log_path)) stop("Missing verified copy log: ", copy_log_path)
fs::dir_create(report_dir, recurse = TRUE)

tibble::tibble(
  setting = c("Search root", "Manifest", "Report folder"),
  value = c(search_root, manifest_path, report_dir)
)
```

## 2. Load and verify the approved staging records

```{r}
#| label: verified-manifest

manifest <- readr::read_csv(manifest_path, show_col_types = FALSE, progress = FALSE)
copy_log <- readr::read_csv(copy_log_path, show_col_types = FALSE, progress = FALSE)

required_manifest <- c(
  "classification", "source_path", "source_relative_path", "source_file",
  "proposed_destination", "reference_file"
)
missing_manifest <- setdiff(required_manifest, names(manifest))
if (length(missing_manifest)) stop("Manifest is missing: ", paste(missing_manifest, collapse = ", "))
if (!all(copy_log$verified_copy)) stop("At least one staged copy is not hash-verified; stop cleanup planning")

top_project <- function(path) {
  relative <- fs::path_rel(fs::path_abs(path), search_root)
  strsplit(as.character(relative), "[/\\\\]")[[1]][1]
}

manifest <- manifest |>
  dplyr::mutate(
    source_project = vapply(source_path, top_project, character(1)),
    source_exists_now = fs::file_exists(source_path),
    staged_copy_exists_now = fs::file_exists(proposed_destination),
    staged_copy_verified = source_path %in% copy_log$source_path[copy_log$verified_copy],
    is_separate_non_snapshot_dataset = source_file %in% separate_dataset_filenames
  )

manifest |>
  dplyr::count(classification, source_exists_now, staged_copy_exists_now, staged_copy_verified)
```

## 3. Enumerate source and configuration files

Generated HTML, JavaScript, CSS, package libraries, caches, the recovery repository, and staged data folders are excluded. The scan targets editable source and configuration files where a data dependency can be changed deliberately.

```{r}
#| label: source-files

all_code <- fs::dir_ls(search_root, recurse = TRUE, type = "file", all = TRUE, fail = FALSE)
relative_code <- fs::path_rel(all_code, search_root)
parts <- stringr::str_split(relative_code, "[/\\\\]")

is_excluded <- vapply(parts, function(x) {
  any(tolower(x) %in% tolower(excluded_directory_names))
}, logical(1))

outside_recovery_repo <- !startsWith(fs::path_abs(all_code), paste0(recovery_root, "/")) &
  !grepl("[/\\\\]SCUC Data Recovery - Superseded[/\\\\]", all_code)

code_files <- all_code[
  !is_excluded & outside_recovery_repo &
    tolower(fs::path_ext(all_code)) %in% source_extensions
]

code_inventory <- tibble::tibble(
  code_path = fs::path_abs(code_files),
  code_relative_path = fs::path_rel(code_files, search_root),
  code_project = vapply(code_files, top_project, character(1)),
  extension = tolower(fs::path_ext(code_files)),
  size_bytes = as.numeric(fs::file_info(code_files)$size)
) |>
  dplyr::filter(size_bytes <= 10 * 1024^2)

code_inventory |>
  dplyr::count(extension, sort = TRUE)
```

## 4. Scan for dependencies

```{r}
#| label: dependency-scan

read_text_lines <- function(path) {
  tryCatch(
    readLines(path, warn = FALSE, encoding = "UTF-8"),
    error = function(e) character()
  )
}

dependency_hits <- dplyr::bind_rows(lapply(seq_len(nrow(code_inventory)), function(i) {
  code <- code_inventory[i, ]
  lines <- read_text_lines(code$code_path)
  if (!length(lines)) return(tibble::tibble())

  dplyr::bind_rows(lapply(seq_len(nrow(manifest)), function(j) {
    item <- manifest[j, ]
    absolute_hit <- grepl(item$source_path, lines, fixed = TRUE)
    relative_hit <- grepl(item$source_relative_path, lines, fixed = TRUE)
    filename_hit <- grepl(item$source_file, lines, fixed = TRUE)
    hit_lines <- which(absolute_hit | relative_hit | filename_hit)
    if (!length(hit_lines)) return(tibble::tibble())

    confidence <- dplyr::case_when(
      any(absolute_hit[hit_lines] | relative_hit[hit_lines]) ~ "certain",
      as.character(code$code_project) == as.character(item$source_project) ~ "likely",
      TRUE ~ "ambiguous"
    )
    match_basis <- dplyr::case_when(
      any(absolute_hit[hit_lines]) ~ "absolute source path",
      any(relative_hit[hit_lines]) ~ "relative source path",
      TRUE ~ "filename only"
    )

    tibble::tibble(
      source_path = item$source_path,
      source_file = item$source_file,
      source_project = item$source_project,
      classification = item$classification,
      reference_file = item$reference_file,
      code_path = code$code_path,
      code_relative_path = code$code_relative_path,
      code_project = code$code_project,
      dependency_confidence = confidence,
      match_basis = match_basis,
      line_number = hit_lines,
      line_excerpt = substr(trimws(lines[hit_lines]), 1L, 500L)
    )
  }))
})) |>
  dplyr::distinct()

readr::write_csv(
  dependency_hits,
  fs::path(report_dir, "scuc_dependency_hits.csv"),
  na = ""
)

dependency_hits |>
  dplyr::count(dependency_confidence, match_basis, sort = TRUE)
```

## 5. Recommend canonical targets

For exact canonical matches, the manifest supplies the authoritative target. For pre-canonical and earlier files, the suggested target is inferred from year coverage and wide/long form. An inferred target is a review aid, not proof of interchangeability.

```{r}
#| label: targets

canonical_target <- function(classification, reference_file, source_file) {
  if (classification == "06 canonical data set matches" && !is.na(reference_file)) {
    return(fs::path(recovery_root, "01 reference canonical", reference_file))
  }
  if (classification == "08 datasets prior to pre-canonical set matches") {
    return(NA_character_)
  }
  text <- tolower(source_file)
  year <- if (grepl("2023", text)) "2023" else "2022"
  form <- if (grepl("wide", text)) "WIDE" else "LONG"
  target_name <- paste0("SCUC_Snapshots_1995_to_", year, "_", form, "_CANONICAL.csv")
  fs::path(recovery_root, "01 reference canonical", target_name)
}

manifest <- manifest |>
  dplyr::mutate(
    recommended_canonical_target = mapply(
      canonical_target, classification, reference_file, source_file,
      USE.NAMES = FALSE
    ),
    canonical_target_exists = !is.na(recommended_canonical_target) &
      fs::file_exists(recommended_canonical_target)
  )
```

## 6. Create the cleanup plan

Only `certain` and `likely` hits block archival. Ambiguous filename-only hits are reported for review but do not by themselves prove that a particular original is in use.

```{r}
#| label: cleanup-plan

dependency_summary <- dependency_hits |>
  dplyr::group_by(source_path) |>
  dplyr::summarise(
    certain_hits = sum(dependency_confidence == "certain"),
    likely_hits = sum(dependency_confidence == "likely"),
    ambiguous_hits = sum(dependency_confidence == "ambiguous"),
    active_code_files = dplyr::n_distinct(code_path[dependency_confidence %in% c("certain", "likely")]),
    active_projects = paste(sort(unique(code_project[dependency_confidence %in% c("certain", "likely")])),
                            collapse = "; "),
    .groups = "drop"
  )

cleanup_plan <- manifest |>
  dplyr::left_join(dependency_summary, by = "source_path") |>
  dplyr::mutate(
    dplyr::across(c(certain_hits, likely_hits, ambiguous_hits, active_code_files),
                  ~ tidyr::replace_na(.x, 0L)),
    active_projects = tidyr::replace_na(active_projects, ""),
    has_active_dependency = certain_hits + likely_hits > 0L,
    cleanup_status = dplyr::case_when(
      is_separate_non_snapshot_dataset ~
        "RETAIN: owner confirmed this is a separate dataset outside the canonical Snapshot cleanup",
      classification == "06 canonical data set matches" & has_active_dependency ~
        "BLOCKED: update code to canonical target, test project, then archive original",
      classification == "06 canonical data set matches" ~
        "READY FOR ARCHIVE after backup approval; exact canonical replacement exists",
      classification == "07 pre-canonical set matches" & has_active_dependency ~
        "BLOCKED: update active code to canonical target; retain documented predecessor in folder 02",
      classification == "07 pre-canonical set matches" ~
        "READY FOR ARCHIVE after backup approval; reference predecessor remains in folder 02",
      classification == "08 datasets prior to pre-canonical set matches" & has_active_dependency ~
        "RETAIN AND INVESTIGATE: active dependency exists; do not archive or delete",
      TRUE ~
        "ARCHIVE CANDIDATE: unmatched historical variant; manual approval required; do not delete"
    ),
    proposed_action = dplyr::case_when(
      is_separate_non_snapshot_dataset ~
        "retain as a separate dataset; no canonical Snapshot replacement required",
      classification == "08 datasets prior to pre-canonical set matches" & has_active_dependency ~
        "retain and investigate; no canonical replacement assumed",
      has_active_dependency ~ "replace/test before archive",
      classification %in% c("06 canonical data set matches", "07 pre-canonical set matches") ~
        "archive original after backup approval",
      TRUE ~ "archive for historical review; no deletion authorized"
    ),
    deletion_authorized = FALSE,
    owner_decision = "pending",
    owner_note = ""
  ) |>
  dplyr::select(
    classification, source_path, source_file, source_project, source_exists_now,
    is_separate_non_snapshot_dataset,
    proposed_destination, staged_copy_exists_now, staged_copy_verified,
    recommended_canonical_target, canonical_target_exists,
    certain_hits, likely_hits, ambiguous_hits, active_code_files, active_projects,
    has_active_dependency, cleanup_status, proposed_action,
    deletion_authorized, owner_decision, owner_note
  ) |>
  dplyr::arrange(dplyr::desc(has_active_dependency), classification, source_project, source_file)

readr::write_csv(cleanup_plan, fs::path(report_dir, "scuc_cleanup_plan.csv"), na = "")

if (requireNamespace("openxlsx", quietly = TRUE)) {
  workbook <- openxlsx::createWorkbook()
  openxlsx::addWorksheet(workbook, "Cleanup plan")
  openxlsx::writeData(workbook, "Cleanup plan", cleanup_plan)
  openxlsx::freezePane(workbook, "Cleanup plan", firstRow = TRUE)
  openxlsx::setColWidths(workbook, "Cleanup plan", cols = 1:ncol(cleanup_plan), widths = "auto")
  openxlsx::addWorksheet(workbook, "Dependency hits")
  openxlsx::writeData(workbook, "Dependency hits", dependency_hits)
  openxlsx::freezePane(workbook, "Dependency hits", firstRow = TRUE)
  openxlsx::saveWorkbook(
    workbook, fs::path(report_dir, "scuc_dependency_scan_and_cleanup_plan.xlsx"), overwrite = TRUE
  )
}

cleanup_plan |>
  dplyr::count(classification, has_active_dependency, cleanup_status, sort = TRUE)
```

## 7. Safe execution order

```{r}
#| label: execution-queue

replacement_queue <- cleanup_plan |>
  dplyr::filter(source_exists_now, has_active_dependency, !is_separate_non_snapshot_dataset) |>
  dplyr::arrange(classification, source_project, source_file)

separate_dataset_retention_queue <- cleanup_plan |>
  dplyr::filter(is_separate_non_snapshot_dataset) |>
  dplyr::arrange(source_project, source_file)

archive_queue <- cleanup_plan |>
  dplyr::filter(source_exists_now, !has_active_dependency, !is_separate_non_snapshot_dataset) |>
  dplyr::arrange(classification, source_project, source_file)

completed_cleanup <- cleanup_plan |>
  dplyr::filter(!source_exists_now, staged_copy_verified, !is_separate_non_snapshot_dataset) |>
  dplyr::arrange(classification, source_project, source_file)

readr::write_csv(replacement_queue, fs::path(report_dir, "scuc_code_replacement_queue.csv"), na = "")
readr::write_csv(archive_queue, fs::path(report_dir, "scuc_archive_candidate_queue.csv"), na = "")
readr::write_csv(completed_cleanup, fs::path(report_dir, "scuc_completed_cleanup.csv"), na = "")
readr::write_csv(
  separate_dataset_retention_queue,
  fs::path(report_dir, "scuc_separate_dataset_retention_queue.csv"),
  na = ""
)

replacement_queue
```

## Interpretation

1. Work through the code-replacement queue first.
2. For each affected project, change the data path to the recommended canonical target and test that project.
3. Re-run this dependency scan; the old file should then have no certain or likely hits.
4. Archive only after a backup exists and the staged copy remains verified.
5. No deletion is authorized by this workbook. Deletion requires a separate reviewed plan.
